---
output:
  html_document: default
  pdf_document: default
---
# What Makes High Quality White Wine?
by Xinkun Chu
========================================================

# Introduction
This analysis use the white wine datasets, which include objective tests (e.g. 
PH values) and sensory data (grade by wine experts), to study the features of 
high graded white wine.  


The dataset is from:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
library(ggplot2)
library(GGally)
library(corrplot)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityWhites.csv")
```

# Univariate Plots Section

A quick look at the dimension and structure of the data.
```{r echo=FALSE}
str(wine)
```
```{r echo=FALSE}
summary(wine)
```


The dataset has 4898 observations with 13 variables. Among the 13 variables, 
1 id. number, 11 physicochemical (fixed acidity, volatile acidity, citric acid,
residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, 
pH, sulphates and alcohol) and 1 sensory (quality, score between 0 and 10).


The main feature is the wine quality graded by the experts between 0 (very bad) 
and 10 (very excellent). The task of this analysis to study which fearures 
affect the white wine quality. The quality table and distribution:

```{r echo=FALSE}
table(wine$quality) 
```

```{r echo=FALSE, fig.height= 4, fig.width= 6}
ggplot(aes(x=quality), data = wine, binwidth=1)+geom_bar()
```

The variance of the quality distribution:
```{r echo=FALSE}
var(wine$quality) 
```


The distribution is slightly left skewed with 6 as the mode. The mean is 5.88, 
and the variance is 0.78. Few are graded lower than 4 or greater than 8.


In this analysis, all the 11 physicochemical features may correlate with the 
quality. Before the correlation study, check the distribution of these features 
first.

```{r, echo=FALSE, warning=FALSE, fig.height=12, fig.width=12}
par(mfrow=c(3,4))
p1 <- boxplot(wine$fixed.acidity, ylab="fixed acidity")
p2 <- boxplot(wine$volatile.acidity, ylab="volatile acidity")
p3 <- boxplot(wine$citric.acid, ylab="citric acid")
p4 <- boxplot(wine$residual.sugar, ylab="residual sugar")
p5 <- boxplot(wine$chlorides, ylab="chlorides")
p6 <- boxplot(wine$free.sulfur.dioxide, ylab="free sulfur dioxide")
p7 <- boxplot(wine$total.sulfur.dioxide, ylab="total sulfur dioxide")
p8 <- boxplot(wine$density, ylab="density")
p9 <- boxplot(wine$pH, ylab="pH")
p10 <- boxplot(wine$sulphates, ylab="sulphates")
p11 <- boxplot(wine$alcohol, ylab= "alcohol")
```


For the 11 features, most of them have outliers, as seen from the boxplot. 
These outliers will be excluded in the histogram distribution. However, the 
cases with grade lower than 4 or greater than 8 is also rare. It's possible 
these outlies contibute to very low or very high grade. These outliers will be 
checked in bivariate analysis. The outliers of some feature will be retained if 
they converge at the two ends of quality, deleted otherwise. 


```{r, echo=FALSE, warning=FALSE, fig.height=12, fig.width=12}
# The distribution of the 11 features. 
# Bin width are adjusted for better display
p1 <- ggplot(data =wine, aes(x=fixed.acidity)) + 
  geom_histogram(binwidth=0.5)+xlim(3,11)
p2 <- ggplot(data =wine, aes(x=volatile.acidity)) + 
  geom_histogram(binwidth=0.03)+xlim(0,0.9)
p3 <- ggplot(data =wine, aes(x=citric.acid)) + 
  geom_histogram(binwidth=0.02)+xlim(-0.01,0.9)
p4 <- ggplot(data =wine, aes(x=residual.sugar)) + 
  geom_histogram(binwidth=1)+xlim(0,28)
p5 <- ggplot(data =wine, aes(x=chlorides)) + 
  geom_histogram(binwidth=0.01)+xlim(0,0.2)
p6 <- ggplot(data =wine, aes(x=free.sulfur.dioxide)) + 
  geom_histogram(binwidth=5)+xlim(0,150)
p7 <- ggplot(data =wine, aes(x=total.sulfur.dioxide)) + 
  geom_histogram(binwidth=10)+xlim(0,300)
p8 <- ggplot(data =wine, aes(x=density)) + 
  geom_histogram(binwidth=0.001)+xlim(0.985,1.005)
p9 <- ggplot(data =wine, aes(x=pH)) + 
  geom_histogram(binwidth=0.1)
p10 <- ggplot(data =wine, aes(x=sulphates)) + 
  geom_histogram(binwidth=0.05)
p11 <- ggplot(data =wine, aes(x=alcohol)) + 
  geom_histogram(binwidth=0.4)

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11, ncol=4)
```

Most of the features have a left skewed distribution. For the citric acid, 
besides the peak around 0.3, there are unusual single value peaks at 0.49 
and 0.74 (g/dm3).  The residual sugar distribution is highly skewed with peak 
around 1~2 g/L, which is expected for the kind of dry wine. We can see the 
distribution better by using transforming the x axis as log10.



```{r, echo=FALSE, fig.width=6, fig.height=4}
ggplot(data =wine, aes(x=residual.sugar)) +
  geom_histogram(binwidth=0.1)+scale_x_log10()
```


```{r echo=FALSE}
# add factor type for quality
# int type is still needed in corrplot
wine$quality.fac <- factor(wine$quality)
```

# Bivariate Plots Section

Check the correlation between these variables.
```{r echo=FALSE, warning=FALSE, fig.width=12, fig.height=12}
corrplot.mixed(cor(wine[2:13]))
```


The residual sugar has a strong correlation of 0.84 with density, and the 
density has a strong negative correlation of -0.78 with alcohol. It seems as
more sugar transformed into alcohol, the alcohol level increases. Meanwhile, 
more sugar transformed means less residual sugar, which leads to a lower 
density of the wine.

For the feature quality, it has a moderate positive correlation of 0.44 with
alcohol, week negative correlation of -0.31 with density, -0.21 with chlorides,
-0.19 with volatile acidity, -0.17 with total sulfur dioxide. The relationship 
between these features and quality will be checked.

```{r,echo=FALSE, warning=FALSE, fig.height=4}
qplot(x = quality.fac, y = alcohol, data = wine, 
      geom = "boxplot", xlab = "quality")
```

There is a tendency that higher quality wine has higher level of alcohol. 


```{r,echo=FALSE, warning=FALSE, fig.height=4}
qplot(x = quality.fac, y = density, data = wine, 
      geom = "boxplot", ylim = c(0.985,1.005),xlab = "quality")
```

High quality wine tends to have lower density. This agrees with result from 
previous plot, because density has a strong negative correlation of -0.78 with 
alcohol. 


```{r,echo=FALSE, warning=FALSE, fig.height=4}
qplot(x = quality.fac, y = chlorides, data = wine,
      geom = "boxplot", ylim = c(0.01,0.1), xlab ="quality")
```

Higher quality wine tends to have a lower level of chlorides, and less spread 
in distribution.


```{r, echo=FALSE, warning=FALSE, fig.height=4}
qplot(x = quality.fac, y = volatile.acidity, data = wine, 
      geom = "boxplot", xlab="quality")
```

No clear trend is seen between volatile acidity and quality from this plot.


```{r, echo=FALSE, warning=FALSE, fig.height=4}
qplot(x = quality.fac, y = total.sulfur.dioxide, data = wine, 
      geom = "boxplot", xlab="quality")
```

No clear trend is seen between total sulfur dioxide and quality from this plot.


# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE}
#re-define the function to be used in this section
mggplot <- function(arg1, arg2){
   p1<- ggplot(aes_string(x = arg1, y = arg2), 
        data = wine, xlab=substitute(arg1)) + 
        geom_point(alpha=0.5, size =2, aes(color = quality.fac)) +
        scale_color_brewer(palette = "YlOrRd", 
                guide = guide_legend(title = 'Quality'))+ 
        scale_x_continuous(lim=c(quantile(get(arg1), 0.01),   
                           quantile(get(arg1), 0.95))) 
   return(p1)
}
```

```{r echo=FALSE, warning=FALSE}
#store the variables to be studied with the same name for convenience
alcohol <- wine$alcohol
density <- wine$density
chlorides <- wine$chlorides
volatile.acidity <- wine$volatile.acidity
total.sulfur.dioxide <- wine$total.sulfur.dioxide
```

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=5}
mggplot("density","alcohol")
```

From the plot, high quality white wine tends to have higher alcohol and lower
density. However, alcohol and density are strongly negative correlated, which
explains for the result.  

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=5}
mggplot("chlorides", "alcohol") 
```

From the plot, high quality white wine tends to have higher alcohol and lower 
chlorides. 

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=5}
mggplot("volatile.acidity", "alcohol") 
```

From the plot, high quality white wine tends to have higher alcohol and lower 
volatile acidity. 

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=5}
mggplot("total.sulfur.dioxide","alcohol")
```


From the plot, higher quality wine tends to have a lower level of total sulfur 
dioxide and higher alcohol. 



------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x=quality), data = wine, binwidth=1)+
  geom_bar(fill=I('#391FF7'))+
  ggtitle("Frequency Distribution of Quality")
```

### Description One

The distribution is slightly left skewed. Most of the white wines are graded 
with 5 or 6. Few are graded lower than 4 or greater than 8. Among the 4898 
observations, only 5 observation for grade 9 and 20 observation for grade 3. 
Because the quality is determined by the median of grade by at least 3 wine 
experts, it's likely that different experts have different opinion on judging 
the quality of the win. So it makes sense that extremely high or low values are 
rare.


### Plot Two
```{r echo=FALSE, Plot_Two}
qplot(x = quality.fac, y = alcohol, data = wine, geom = "boxplot", 
      xlab = "quality ",ylab="alcohol (%)")+
  ggtitle("alcohol level by quality")
```

### Description Two

Alcohol has a moderate positive correlation of 0.44 with quality, which is the
largest correlation with quality found among the features. The boxplot shows 
the alcohol percentage distribution for each quality. There is a tendency that
higher quality wine has higher level of alcohol. Of course, more alcohol 
doesn't necessarily mean higher quality wine. Many factors may explain this
tendency. It's possible that higher alcohol are the result of better quality 
grapes and better fermentation. 

### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three,fig.width=8, fig.height=5}
mggplot("total.sulfur.dioxide","alcohol")+
  labs(title="Scatter plot of alcohol and total sulfur dioxideby quality",
       x = "Total sulfur dioxide (mg/dm3)",y = "Alcohol (%)")
```

### Description Three

In the scatter plot of total sulfur dioxide and alcoholm, the quality are 
differentiate by color. Most of the light circles(wine with quality 5)
converge at the right bottom of the plot. Low quality wine tends to have a 
higher level of total sulfur dioxide and lower alcohol. While most of the 
dark circles (quality 7,8) have higher level of alcohol 
and total sulfur dioxide less than 150 mg/dm3.


------

# Reflection
In this analysis, the correlation between white wine quality and 11 
physicochemical features are studied using a dataset of 4898 observations. 
It's found that alcohol has a moderate positive correlation with quality, 
while density, chloride, volative acidity and total sulfur dioxide have low 
negative correlation with wine quality. However, it's hard to tell what 
physicochemical features a good white wine should have. There is no obvious
liner relationship between these factors, further study may need the tools of 
machine learning to train a model to dig out the complicated relationships. 
Also, the samples for high(>8) and low quality(<4) wine would be important for 
study, but the statistics is not enough in the dataset.

